---
title: "insectai_to_camtrapdp"
author: 
- "Sanne Govaert"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages

Before beginning, you need to ensure that all required R packages are installed. This includes packages for data handling, working with frictionless data, and the development version of `camtrapdp` from GitHub.

```{r}
# Install required packages if not already installed
rlang::check_installed(c(
  "here",
  "dplyr",
  "frictionless",
  "camtrapdp"
))
```


## Load libraries

Next, load all the necessary libraries.

```{r}
library(here)
library(dplyr)
library(frictionless)
library(camtrapdp)
```

# Import data

Import the raw acoustic data from a CSV file. This file contains the source data that will be transformed into a Camtrap DP.

```{r}
project_name <- "my_project"
file_name <- "my_file.csv"
data <- read.csv(here::here("data", "raw", project_name, file_name))
```

# Create resources

Create the five resources (deployments, media, observations, detections, models) based on the raw data. You can use the code below as a starting point, but you will need to fill in the missing information (currently all set to `NA`).

## Create deployments resource

```{r}
deployments <-
  data |> 
  dplyr::mutate(
    deploymentID = NA_character_, # Required
    locationID = NA_character_,
    locationName = NA_character_,
    latitude = NA_real_, # Required
    longitude = NA_real_, # Required
    coordinateUncertainty = NA_integer_,
    deploymentStart = NA_character_, # Required | YYYY-MM-DDThh:mm:ssZ or YYYY-MM-DDThh:mm:ss±hh:mm
    deploymentEnd = NA_character_, # Required | YYYY-MM-DDThh:mm:ssZ or YYYY-MM-DDThh:mm:ss±hh:mm
    setupBy = NA_character_,
    cameraID = NA_character_,
    cameraModel = NA_character_,
    cameraDelay = NA_integer_,
    cameraHeight = NA_real_,
    cameraDepth = NA_real_,
    cameraTilt = NA_integer_,
    cameraHeading = NA_integer_,
    detectionDistance = NA_real_,
    timestampIssues = NA_character_,
    attractantUse = NA,
    atttractantTypr = NA_character_,
    featureType = NA_character_,
    habitat = NA_character_,
    deploymentGroups = NA_character_,
    deploymentTags = NA_character_,
    deploymentComments = NA_character_,
    .keep = "none"
  ) |> 
  dplyr::arrange(deploymentID)
```

## Create media resource

```{r}
media <-
  data |> 
  dplyr::mutate(
    mediaID = NA_character_, # Required
    deploymentID = NA_character_, # Required
    captureMethod = NA_character_,
    timestamp = NA_character_, # Required | YYYY-MM-DDThh:mm:ssZ or YYYY-MM-DDThh:mm:ss±hh:mm
    filePath = NA_character_, # Required
    filePublic = NA, # Required
    fileName = NA_character_,
    fileMediatype = NA_character_, # Required
    exifData = NA_character_,
    favorite = NA_character_,
    mediaComments = NA_character_,
    .keep = "none"
  ) |> 
  dplyr::arrange(mediaID) |>
  dplyr::distinct()
```

## Create observations resource

```{r}
observations <-
  data |> 
  dplyr::mutate(
    observationID = NA_character_, # Required
    deploymentID = NA_character_, # Required
    mediaID = NA_character_,
    eventID = NA, # If empty, set to NA to be able to use read_camtrapdp(), see https://github.com/inbo/camtrapdp/issues/162
    eventStart = NA_character_, # Required
    eventEnd = NA_character_, # Required
    observationLevel = NA_character_, # Required
    observationType = NA_character_, # Required 
    cameraSetupType = NA_character_,
    scientificName = NA_character_,  # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    classificationMethod = NA_character_,
    classifiedBy = NA_character_,
    classificationTimestamp = NA_character_,
    classificationProbability = NA_real_,
    observationTags = NA_character_,
    observationComments = NA_character_,
    .keep = "none"
  ) |> 
  dplyr::arrange(observationID)
```

## Create detections resource

```{r}
detections <-
  data |> 
  dplyr::mutate(
    detectionID = NA_character_, # Required
    detectionsourceID = NA_character_,
    observationID = NA_character_,
    deploymentID = NA_character_, # Required
    mediaID = NA_character_,
    eventID = NA,
    eventStart = NA_character_, # Required
    eventEnd = NA_character_, # Required
    detectionLevel = NA_character_, # Required
    detectionType = NA_character_, # Required 
    cameraSetupType = NA_character_,
    scientificName = NA_character_,  # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    classificationMethod = NA_character_,
    classifiedBy = NA_character_,
    modelID = NA_character_, 
    classificationTimestamp = NA_character_,
    classificationProbability = NA_real_,
    detectionsTags = NA_character_,
    detectionComments = NA_character_,
    .keep = "none"
  ) |> 
  dplyr::arrange(detectionID)
```

## Create models resource

```{r}
models <-
  data |>
  dplyr::mutate(
    modelID = NA_character_, # Required
    modelName = NA_character_, 
    modelVersion = NA_character_,
    modelDescription = NA_character_,
    modelPathRpository = NA_character_, # Required
    publicationDate = NA_character_,
    modelTask = NA_character_, # format": "%Y-%m-%dT%H:%M:%S%z
    .keep = "none"
  ) |>
  dplyr::arrange(modelID)
```


# Create data package

A Camtrap DP is a Frictionless Data Package that consists of
- datapackage.json: Metadata about the data package and camera trap project.
- deployments.csv: Table with camera trap placements (deployments).
- media.csv: Table with media files recorded during deployments.
- observations.csv: Table with observations derived from the media files.

For the insectai adaptation, two more resources are added:
- detecions.csv
- models.csv

Start with compiling the five resources into a frictionless data package, adding metadata such as schema files and project information.

```{r}
insectai <-
  frictionless::create_package() |>
  frictionless::add_resource(resource_name = "deployments", data = deployments) |>
  frictionless::add_resource(resource_name = "media", data = media) |>
  frictionless::add_resource(resource_name = "observations", data = observations) |>
  frictionless::add_resource(resource_name = "detections", data = detections) |>
  frictionless::add_resource(resource_name = "models", data = models) |>
  append(c(
    name = "camtrap-dp-insectai-example-dataset", # Required
    profile = "https://raw.githubusercontent.com/tdwg/camtrap-dp/refs/heads/insectai/camtrap-dp-profile.json", # Required
    id = NULL,
    created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    title = NULL,
    description = NULL,
    version = "1.0",
    keywords = NULL,
    image = NULL,
    homepage = NULL,
    sources = NULL,
    bibliographicCitation = NULL,
    coordinatePrecision = NULL,
    spatial = NULL, # Required, but keep this empty, this comes later
    temporal = NULL, # Required, but keep this empty, this comes later
    taxonomic = NULL, # Required, but keep this empty, this comes later
    relatedIdentifiers = NULL,
    references = NULL
  )) |>
  frictionless::create_package()
```

Then add contributor information, licenses, and project metadata to complete the package description.

```{r}
insectai$contributors <- list( # Required
  list(
    title = NULL,
    email = NULL,
    path = NULL,
    role = NULL,
    organization = NULL
  )
)

insectai$licenses <- list(
  list(
    name = "CC0-1.0",
    scope = "data" # Required
  ),
  list(
    path = "http://creativecommons.org/licenses/by/4.0/",
    scope = "media" # Required
  )
)

insectai$project <- list( # Required
  # id = "",
  title = NULL, # Required
  # acronym = "",
  description = NULL,
  protocolType = NULL, # Required
  samplingDesign = NULL, # Required
  # path = NULL,
  captureMethod = NULL, # Required
  individualAnimals = NULL, # Required
  observationLevel = NULL # Required
)
```

## Add taxonomic coverage

Create a function to format taxonomy rows for inclusion in the metadata, then apply it to the data.
The code is to simply add all scientific names.
You can also add the whole taxonomy and vernacular names to the metadata, see https://camtrap-dp.tdwg.org/metadata/#taxonomic.

```{r}
taxonomyy <- data |> 
  dplyr::select(scientificName) |>
  dplyr::distinct()

map_taxonomy <- function(taxonomy) {
  list(
    scientificName = taxonomy$scientificName,
    taxonRank = "species"
  )
}
```

```{r}
taxonomy_list <- apply(taxonomy, 1, as.list)
insectai$taxonomic <- purrr::map(taxonomy_list, map_taxonomy)
```

Or you can also simplify it by only providing higher taxonomic levels, eg:

```{r}
taxonomic_coverage <-
  data.frame(
    scientificName = "Insecta",
    taxonRank = "class"
    )
insectai$taxonomic <- taxonomic_coverage
```

# Write Data Package

## Create frictionless data package

Save the frictionless data package, including all resources and metadata, to the designated interim directory.

```{r}
frictionless::write_package(insectai, here::here("data", "interim", project_name))
```

## Read with camtrapdp

Now, read the Camtrap DP data package to automatically update the spatial, temporal, and taxonomic coverage fields in the metadata based on the resources.

```{r}
insectai_dp <- camtrapdp::read_camtrapdp(here::here("data", "interim", project_name, "datapackage.json"))
```

## Write Camtrap DP

Finally, we write the completed Camera Trap Data Package to the processed directory.

```{r}
camtrapdp::write_camtrapdp(insectai_dp, here::here("data", "processed", project_name))
```

This final DP can be validated against the specifications of Camtrap DP (and Frictionless Data Package). See https://camtrap-dp.tdwg.org/#validation.
